---
title: "Publishable, reproducible map of sample sites"
output: html_notebook
---

# Background

Today we are going to use R to aggregate geospatial data of several different kinds and make a publishable and reproducible map. The demonstration map has been adapted from one that I made for a publication a few years ago (https://academic.oup.com/auk/article/138/2/ukaa086/6167858). In this project, we examined the effects of sound broadcast treatments on settlement decisions of endangered Marbled Murrelets. Note that because the species is endangered, I have scrambled the coordinates of our study sites (i.e., they're not the real sites we used). The map we are producing is going to have 3 elements:

1. A zoomed-in view of our study area;
2. A zoomed-out view of the stat of Oregon with an extent rectangle for the study area; and
3. A map legend.

# Prep work

Let's start by loading the necessary packages for this exercise and bringing in a raw dataset of the sample points.
```{r}
require(tidyverse)


read.csv()
```

And this chunk creates
```{r}



ggplot(site, aes(x=easting, y=northing))+
  geom_point()
```

